Towards a miniKanren with fair search strategies

### 2019-05-22 Paper/paper.tex
lots of corrections

%% scheme-list :
\usepackage{listings}
\usepackage{color}
\usepackage{textcomp}
\usepackage{trackchanges}

@@

\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
  		   {repeato}{{repeat$^o$}}{6}
}
%% scheme-list .

@@

\newcommand{\conde}{\texttt{cond$^e$}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi }[0]{DFS$_\textrm{i}$}
\newcommand{\DFSf }[0]{DFS$_\textrm{f}$}
\newcommand{\DFSbi}[0]{DFS$_\textrm{bi}$}
\newcommand{\BFS  }[0]{BFS}

\newcommand{\BFSopt}[0]{BFS$_\textrm{opt}$}

\newcommand{\BFSeff}[0]{BFS$_\textrm{eff}$}

\newcommand{\BFSser}[0]{BFS$_\textrm{ser}$}
\newtheorem{defn}{Definition}[section]

@@

  We describe fairness levels in disjunction and conjunction
  implementations.  Specifically, a disjunction implementation can be
  fair, almost-fair, or unfair. And a conjunction implementation can
  be fair or unfair.  We compare the fairness level of four search
  strategies: the standard miniKanren interleaving depth-first search (\DFSi),
  the balanced interleaving depth-first search (\DFSbi), the fair depth-first
  search (\DFSf), and the standard breadth-first search (\BFSser).  
  \DFSbi{} and \DFSf{} are new. And we present a new, more efficient
  and simpler implementation of \BFSser. Using quantitative evaluation, 
  we argue that
  \DFSbi{} and \DFSf{} are competitive alternatives to \DFSi, 
  and that \BFSopt{} is more efficient than the standard \BFSser.

% The syntax of a programming language should reflect its semantics. When writing 
% a \conde{} expression in miniKanren, a programmer would expect all clauses 
% share the same chance of being explored, as these clauses are written in 
% parallel. The existing search strategy, interleaving depth-first search 
% (\DFSi{}), 
% however, prioritizes its clauses by the order how they are written down. 
% Similarly, when a \conde{} is followed by another goal conjunctively, a 
% programmer would expect states in parallel share the same chance of being 
% explored. Again, the answers by \DFSi{} is different from the 
% expectation. We have devised three new search strategies that have different 
% level of fairness in \disj{} and \conj{}.
% 
% \NOTE[LKC]{\disj{} and \conj{} occurs free in the abstract.}

%The syntax of a programming language should reflect its semantics. When using 
%a disjunctive operator in relational programming, a programmer would expect 
%all 
%sub-goals of this disjunct to share the same chance of being explored, as 
%these 
%clauses are written in parallel. The existing multi-arity disjunctive operator 
%in miniKanren, however, prioritize its clauses by the order of which these 
%clauses are written down. We have devised two new search strategies that 
%allocate computational effort more fairly in all clauses.

@@

miniKanren is a family of relational programming languages.
\citet{friedman_reasoned_2018} introduce miniKanren and its
implementation in \emph{The Reasoned Schemer, 2nd Ed}
(TRS2). miniKanren programs, have been proven to be useful in solving 
many problems as in miniKanren.org.\citet{byrd2017unified}.

@@

\NOTE[DPF]{I think that the abbreviations for the search strategies
should not come up until the Introduction.}

\NOTE[DPF]{We should reference the earliest paper (See Will Byrd vita)}

\NOTE[DPF]{When we talk about complete and incomplete, I think we should
make it explicit that we are only talking about (run n (x y z) g ...) and
not (run* (x y z) g ...), but an alternative is that a stream is \emph{productive}
or \emph{unproductive}.}

\NOTE[LKC]{Shall we delete the sentence about relational interpreters? It is 
misleading: it is emphasizing relational interpreters, but we say nothing 
further about them later.}
\NOTE[DPF]{I have deleted the sentence.}
% \conde{} is the flagship disjunction operator in miniKanren.
A subtlety arises 
when a \conde{} contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is used to generate simple and shallow programs,
the clauses for constructors should be put at the top of the
\conde{} clauses.
When performing proof searches for complicated programs, the clauses for 
eliminators should be put at the top of \conde{} clauses. With \DFSi, these two uses cannot be 
efficient at the same time. In fact, to make one use efficient, the other one 
must be more sluggish.

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause similar priority. 
Fair \conj{}, a related concept, is more subtle. The discussion of fair
\conj{} is covered in the next section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSopt), 
        a search strategy 
	with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSopt{} implementation is semantically equivalent to \BFSser
however, Our code runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints. Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

% Goals check their input states and outputs possibly extended states. A goal 
% might fail for some inputs, in which case the output search space would be 
% empty.

@@

Running the same program again with almost-fair \disj {} (e.g. 
DFS$_\textrm{bi}$) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

@@

\DFSbi{} is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde{} has five clauses, the clauses 
of \texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

@@

\begin{defn}[almost-fair \disj{}]
A \disj{} is almost-fair if and only if it allocates computational resources
evenly to search spaces produced by goals in the same disjunction that 
the maximal ratio of resources is bounded by a constant.
\end{defn}

@@

\subsection{fair \texttt{conj}}

\NOTE[LKC]{Checked grammer of text before this line}

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.

\NOTE[LKC]{Should I add a footnote?}

``
Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to a relational definition 
in \conde{} clauses (including this case).

@@

Under the hood, the \conde{} and the call to \texttt{repeato} are connected by 
\conj{}. The \conde{} goal outputs a search space including four trivially 
different states. Then the next conjunctive goal, \texttt{(repeato x q)}, is 
applied to each of these states, producing four trivially different search 
spaces. In the examples above, the \conj{}s are allocating more computational 
resources to the search space of \texttt{a}. On the contrary, fair \conj{} 
would allocate resources evenly to each search space. For example,

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but there
would not be any lists of length two or longer, which makes the strategies incomplete. 
A search strategy is \emph{complete} if and only if it can determine all the answers within finite 
time, otherwise, it is \emph{incomplete}.

\NOTE[DPF]{Does this assume that we never use run*, since in fact, the paper does not
include and uses of run*.  Something should be mentioned around here.}

@@

\BFSser{} puts states of the same cost in the same bag, and allocates
resources carefully among search spaces related to different bags such
that answers are produced in increasing order of cost. The \emph{cost}
of an answer is its depth in the search tree (i.e., the number of
calls to relations required to find them) \citep{seres1999algebra}. In
the following example, every answer is a list of a list of symbols,
where inner lists are the same. Here the cost of each answer is equal
to the length of its inner list plus the length of its outer list.

@@

% Our definition of fair \conj{} is orthogonal with completeness. For 
% example, a naively fair strategy is fair but not complete, while BFS is fair and 
% complete.

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2,
chapter 10 and the appendix, ``Connecting the wires,'' 
provides a comprehensive description of the 
miniKanren 
implementation but limited to unification constraints ($\equiv$).

@@

\NOTE[DPF]{If we use the name subscript, the subscripts should use an underscore and we should do it for gunderscorel and gundescorer. If we can use an l that
looks less like a 1, that would even be better.  Look at the l in TLT.}

Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search 
strategies. We follow some conventions to name variables: variables bound to 
states are named \texttt{s}; variables bound to goals are named \texttt{g} 
(possibly with subscript); variables bound to search spaces have names ending 
with $^\infty$. Fig.~\ref{DFSi-0} shows the implementation of \disj. The 
first function, \texttt{disj2}, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting search spaces with \texttt{append$^\infty$}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \texttt{conj2}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The latter process is done with a 
the helper function \texttt{append-map$^\infty$} that applies its input goal to states 
in its input search spaces and composes the resulting search spaces. It reuses 
\texttt{append$^\infty$} for search space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has an almost-fair \disj{} and unfair 
\conj{}. The implementation of DFS$_\textrm{bi}$ differs from 
DFS$_\textrm{i}$'s in the \disj{} macro. We list the new \disj{} with its 
helper in Fig.~\ref{balanced-disj}. When there are one or more disjunctive 
goals, \disj{} builds a balanced binary tree whose leaves are the goals and 
whose nodes are \texttt{disj2}s, hence the name of this search strategy. 
The new helper, \texttt{disj+}, takes two additional `arguments'. They 
accumulate goals to be put in the left and right subtrees. The first clause 
handles the case where there is only one goal. In this case, the tree is the 
goal itself. When there are more goals, we partition the list of goals 
into two sublists of roughly equal lengths and recur on the two sublists. Goals 
are moved to the accumulators in the last clause. As we are moving 
two goals each time, there are two base cases: (1) no goal remains; (2) one 
goal remains. These two new base cases are handled in the second clause and the 
third clause, respectively. In contrast, the \disj{} in \DFSi{} constructs the 
binary tree in a particularly unbalanced form.

@@

\NOTE[DPF]{Why does the pair? case precede the null? case? Is it simply more efficient? It shouldn't be}

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is changed to call a new and 
fair version of \texttt{append$^\infty$}. \texttt{append$^\infty_{fair}$} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that 
\texttt{s$^\infty$} and \texttt{t$^\infty$} haven't been swapped. The swapping 
happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s$^\infty$} and 
\texttt{t$^\infty$} are ended with a thunk and they have been swapped. In this 
case, a new thunk is constructed. Here changing the order of 
\texttt{t$^\infty$} and \texttt{s$^\infty$} won't hurt the fairness (though it 
changes the order of answers). We swap them back so that answers are 
produced in a more natural order.

\NOTE[DPF]{When or where do you swap them back?}

@@

\BFSser{} has both fair \disj{} and fair \conj{}. Our implementation is based on 
DFS$_\textrm{f}$ (not DFS$_\textrm{i}$). To implement \BFSopt{} based on \DFSf{}, 
we need \texttt{append-map$^\infty_{fair}$} in addition to 
\texttt{append$^\infty_{fair}$}. 
The only difference between \texttt{append-map$^\infty$} and 
\texttt{append-map$^\infty_{fair}$} is that the latter calls 
\texttt{append$^\infty_{fair}$} instead of \texttt{append$^\infty$}.

The implementation can be improved in two ways. First, as mentioned in 
section 2.2, \BFSser puts answers in bags and answers of the same cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, \texttt{append$^\infty_{fair}$} is extravagant in memory 
usage. It 
makes $O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

Both \BFSser{} \citet{seres1999algebra} and BFSopt{} produce answers in 
increasing order of cost. So it is interesting to see if they are equivalent. 
We prove so in Coq. The details are in the second subsection.

@@

\NOTE[DPF]{Never refer to a literal section like 2.2. Instead
stick a label where it starts right near the section number and
stick another label where the subsection starts.
Then ref the two labels and place a period between them.}.

As mentioned in section 2.2, BFS puts answers in bags and answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are all in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. We make these facts more obvious by 
changing the type of search space, modifying related function definitions, 
and introducing a few more functions.

@@

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). They are compared with 
\citeauthor{seres1999algebra}'s implementation in our proof. The first three 
functions in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} 
makes an empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions as before.

@@

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. The \BFSopt{} implementation is not 
an exception: \texttt{append-map$^\infty_{fair}$} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append$^\infty_{fair}$} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, \texttt{ks} is called with the first answer and the rest of the space. 
Otherwise, \texttt{kf} is called with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is like an eliminator of 
search spaces, hence the name. The remaining functions do the same thing as 
before.

\subsection{comparison with the BFSser and BFSopt}

In this section, we compare the pure subset of \BFSopt{} with the 
\BFSser{}. We focus on the pure subset because \BFSser{} is pure.
and represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate \BFSser{}'s Haskell code into Racket (See supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than \BFSopt{}'s implementation. Details about 
differences in efficiency are in the table below.

We use Coq to show that \BFSser{} and \BFSopt{} are 
semantically equivalent, (i.e., \texttt{(run n ? g)} 
produces the same result (See supplements for the formal proof).

@@

\citet{seres1999algebra} also describe a breadth-first search 
strategy. We prove \BFSser{} is semantically equivalent to \BFSopt. But,
The implementation of \BFSopt is shorter and performs better in comparison with a straightforward translation of their Haskell code.

>>>>

\usepackage{color}
\usepackage{textcomp}
\usepackage{hyperref}

\usepackage{trackchanges}

@@

\usepackage{listings}
\lstset{
  language=Scheme,
  basicstyle=\ttfamily,
  morekeywords={run,conde,run*,defrel,==,fresh,disj,conj,disj+},
  alsodigit={!\$\%&*+-./:<=>?@^_~},
  morecomment=[l]{;,\#{}lang},
  mathescape=true,
  literate={-inf}{{$^\infty$}}{1} {/fair}{{$_{fair}$}}{3}
  		   {→}{{$\rightarrow$}}{1}
  		   {×}{{$\times$}}{1}
  		   {==}{{$\equiv$}}{1}
  		   {conde}{{\textcolor[rgb]{0,.3,.7}{\textbf{cond$^e$}}}}{4}
  		   {repeato}{{repeat$^o$}}{6}
  		   {g0}{{g$_0$}}{2}
  		   {g1}{{g$_1$}}{2}
  		   {g2}{{g$_2$}}{2}
  		   {gl}{{g$_l$}}{2}
  		   {gr}{{g$_r$}}{2}
  		   {DFSi }{{DFS$_i$}}{3}
  		   {DFSbi}{{DFS$_{bi}$}}{3}
  		   {DFSf }{{DFS$_f$}}{3}
  		   {BFSser}{{BFS$_{ser}$}}{3}
}
%% scheme-list .

@@

\newcommand{\conde}{\texttt{cond$^e$}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}

\newcommand{\veryrecursiveo}{\texttt{very-recursive$^o$}}
\newcommand{\appendo}{\texttt{append$^o$}}
\newcommand{\reverso}{\texttt{revers$^o$}}

\newcommand{\DFSi }[0]{DFS$_\textrm{i}$}
\newcommand{\DFSf }[0]{DFS$_\textrm{f}$}
\newcommand{\DFSbi}[0]{DFS$_\textrm{bi}$}
\newcommand{\BFSser}[0]{BFS$_\textrm{ser}$}

\newtheorem{defn}{Definition}[section]

@@

We describe fairness levels in disjunction and conjunction
implementations.  Specifically, a disjunction implementation can be
fair, almost-fair, or unfair. And a conjunction implementation can
be fair or unfair.  We compare the fairness level of four search
strategies: the standard miniKanren interleaving depth-first search (\DFSi),
the balanced interleaving depth-first search (\DFSbi), the fair depth-first
search (\DFSf), and the standard breadth-first search (\BFSser).
\DFSbi{} and \DFSf{} are new. And we present a new, more efficient
and simpler implementation of \BFSser. Using quantitative evaluation, 
we argue that
\DFSbi{} and \DFSf{} are competitive alternatives to \DFSi, 
and that our \BFSser{} implementation is more efficient than the current one.

% The syntax of a programming language should reflect its semantics. When 
%writing 
% a \conde{} expression in miniKanren, a programmer would expect all clauses 
% share the same chance of being explored, as these clauses are written in 
% parallel. The existing search strategy, interleaving depth-first search 
% (\DFSi{}), 
% however, prioritizes its clauses by the order how they are written down. 
% Similarly, when a \conde{} is followed by another goal conjunctively, a 
% programmer would expect states in parallel share the same chance of being 
% explored. Again, the answers by \DFSi{} is different from the 
% expectation. We have devised three new search strategies that have different 
% level of fairness in \disj{} and \conj{}.
%

@@

miniKanren is a family of relational programming languages.
\citep{Friedman:2005:RS:1121583} \citet{friedman_reasoned_2018} introduce 
miniKanren and its
implementation in \emph{The Reason Schemer} 
and \emph{The Reasoned Schemer, 2nd Ed} (TRS2). \citet{byrd2017unified} have 
proved that miniKanren programs are useful in solving many problems as in 
miniKanren.org.

@@

\NOTE[DPF]{I think that the abbreviations for the search strategies
should not come up until the Introduction.}
\NOTE[LKC]{Do you mean they should not come up in Abstract?}

\NOTE[DPF]{When we talk about complete and incomplete, I think we should
make it explicit that we are only talking about (run n (x y z) g ...) and
not (run* (x y z) g ...), but an alternative is that a stream is \emph{productive}
or \emph{unproductive}.}
\NOTE[LKC]{I have added a sentence at the place where we define
completeness.}

% \conde{} is the flagship disjunction operator in miniKanren.
A subtlety arises 
when a \conde{} contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves all
purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When the type checker is generating simple and shallow programs,
the clauses for constructors had better be at the top of the
\conde{} expression.
When performing proof searches for complicated programs, the clauses for 
eliminators had better be at the top of the \conde{} expression. With \DFSi, 
these two uses cannot be efficient at the same time. In fact, to make one use 
efficient, the other one must be more sluggish.

The specification that gives every clause in the same \conde{} equal 
``search priority'' is fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause similar priority. 
Fair \conj{}, a related concept, is more subtle. We cover it in the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (\BFSser), a search 
strategy with fair \disj{} and fair \conj{}. And we prove formally that our 
\BFSser{} implementation is semantically equivalent to \BFSser{}.
Our code runs faster in all benchmarks and we believe it is simpler.
\end{itemize}

@@

In this section, we define fair \disj{}, almost-fair \disj{} and fair \conj{}. 
Before going further into fairness, we give a short review of the terms:
\emph{state}, search \emph{space}, \emph{goal}.
A \emph{state} is a collection of constraints. (Here, we restrict 
constraints to unification constraints.) Every answer corresponds to a 
state. A space is a collection of states. And a \emph{goal} is a 
function from a state to a space.

% Goals check their input states and outpust possibly extended states. A goal 
% might fail for some inputs, in which case the output search space would be 
% empty.

@@

Running the same program again with almost-fair \disj {} (e.g. 
\DFSbi{}) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

@@

\DFSbi{} is fair only when the number of goals is a power of 2, 
otherwise, it allocates some goals twice as many resources as the 
others. In the above example, where the \conde{} has five clauses, \DFSbi{} 
allocates more resources to the clauses of \texttt{b}, \texttt{c}, and 
\texttt{d}.

@@

\begin{defn}[almost-fair \disj{}]
A \disj{} is almost-fair if and only if it allocates computational resources
so evenly to search spaces produced by goals in the same disjunction that 
the maximal ratio of resources is bounded by a constant.
\end{defn}

\NOTE[LKC]{I might be wrong, but `\emph{so} evenly ... that the maximal ...' 
looks more natural to me}

@@

\subsection{fair \texttt{conj}}
\label{sec:fairconj}

\NOTE[LKC]{I change `1/4 in the answer' to `1/4 in the answer list'. }

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer list. Search strategies with unfair \conj{} 
(e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} were to contain more clauses.
Although some strategies have different level of fairness in \disj{}, they 
have the same behavior when there is no call to a relational definition in 
\conde{} clauses (including this case).

@@

Under the hood, the \conde{} and the call to \texttt{repeato} are connected by 
\conj{}. The \conde{} goal outputs a search space including four trivially 
different states. 
Applying the next conjunctive goal, \texttt{(repeato x q)}, produces four 
trivially different search spaces.
In the examples above, all search strategies allocate more computational 
resources to the search space of \texttt{a}. On the contrary, fair \conj{} 
would allocate resources evenly to each search space. For example,

@@

A more interesting situation is when the first conjunct produces an unbounded
number of states. Consider the following example, a naive specification of 
fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but 
there
would not be any lists of length two or longer, which makes the strategies 
incomplete. 
A search strategy is \emph{complete} if and only if ``every correct answer 
would be discovered after some finite time'' \cite{seres1999algebra}, 
otherwise, it is \emph{incomplete}. In the 
context of miniKanren, a search strategy is complete means that every correct 
answer has a positition in large enough answer lists.

\NOTE[DPF]{Does this assume that we never use run*, since in fact, the paper 
does not include and uses of run*.  Something should be mentioned around here.}
\NOTE[LKC]{We never use run* in this paper because fairness is only 
interested when we have an unbounded number of answers. However, it is 
perfectly fine to use run* with any search strategies.}

@@

\BFSser{} puts states of the same cost in the same bag, and allocates
resources carefully among search spaces related to different bags such
that it produces answers in increasing order of cost. The \emph{cost}
of an answer is its depth in the search tree (i.e., the number of
calls to relations required to find them) \citep{seres1999algebra}. In
the following example, every answer is a list of a list of symbols,
where inner lists are the same. Here the cost of each answer is equal
to the length of its inner list plus the length of its outer list.

@@

@@

In this section, we review the implementation of interleaving depth-first 
search (\DFSi). We focus on parts that are relevant to this paper. TRS2,
chapter 10 and the appendix, ``Connecting the wires'', 
provides a comprehensive description of the 
miniKanren 
implementation but limited to unification constraints ($\equiv$).

@@

\NOTE[DPF]{If we use the name subscript, the subscripts should use an 
underscore and we should do it for gunderscorel and gundescorer. If we can use 
an l that looks less like a 1, that would even be better.  Look at the l in 
TLT.}
\NOTE[LKC]{I have turned those `subscripts' to real subscripts.}

Fig.~\ref{DFSi-0} and Fig.~\ref{DFSi-1} show parts that are later compared 
with other search 
strategies. We follow some conventions to name variables: \texttt{s}s name 
states; \texttt{g}s (possibly with subscript) name goals; 
variables ending with $^\infty$ name search spaces. Fig.~\ref{DFSi-0} shows the 
implementation of \disj. The 
first function, \texttt{disj2}, implements binary disjunction. It applies the 
two disjunctive goals to the input state \texttt{s} and composes the two 
resulting search spaces with \texttt{append$^\infty$}. The following syntax 
definitions say \disj{} is right-associative. Fig.~\ref{DFSi-1} 
shows the implementation of \conj. The first function, \texttt{conj2}, 
implements binary conjunction. 
It applies the \emph{first} goal to the input state, then applies the second 
goal to states in the resulting search space. The helper function 
\texttt{append-map$^\infty$} applies its input goal to states 
in its input search spaces and composes the resulting search spaces. It reuses 
\texttt{append$^\infty$} for search space composition. The following syntax 
definitions say \conj{} is also right-associative.

@@

Balanced interleaving DFS (\DFSbi{}) has an almost-fair \disj{} and unfair 
\conj{}. The implementation of \DFSbi{} differs from 
\DFSi{}'s in the \disj{} macro. We list the new \disj{} with its 
helper in Fig.~\ref{balanced-disj}. When there are one or more disjunctive 
goals, \disj{} builds a balanced binary tree whose leaves are the goals and 
whose nodes are \texttt{disj2}s, hence the name of this search strategy. 
The new helper, \texttt{disj+}, takes two additional `arguments'. They 
accumulate goals to be put in the left and right subtrees. The first clause 
handles the case where there is only one goal. In this case, the tree is the 
goal itself. When there are more goals, we partition the list of goals 
into two sublists of roughly equal lengths and recur on the two sublists. We 
move goals to the accumulators in the last clause. As we are moving 
two goals each time, there are two base cases: (1) no goal remains; (2) one 
goal remains. We handle these two new base cases in the second clause and the 
third clause, respectively. In contrast, the \disj{} in \DFSi{} constructs the 
binary tree in a particularly unbalanced form.

@@

\NOTE[DPF]{Why does the pair? case precede the null? case? Is it simply more 
efficient? It shouldn't be}
\NOTE[LKC]{I did for efficiency. I have re-ordered two cases back.}

@@

Fair DFS (\DFSf) has fair \disj{} and unfair \conj{}. The 
implementation of \DFSf{} differs from \DFSi{}'s in 
\texttt{disj2} (Fig.~\ref{fDFS}). The new \texttt{disj2} calls a new and 
fair version of \texttt{append$^\infty$}. \texttt{append$^\infty_{fair}$} 
immediately 
calls 
its helper, \texttt{loop}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that we haven't swapped
\texttt{s$^\infty$} and \texttt{t$^\infty$}. The swapping 
happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s$^\infty$} and 
\texttt{t$^\infty$} are ended with a thunk and we have swapped them. In this 
case, we construct a new thunk. The new thunk swaps two spaces back in the
recursive call. This is unnecessary for fairness. We do it to produce answers 
in a more natural order.

\NOTE[DPF]{When or where do you swap them back?}
\NOTE[LKC]{In the recursion}

@@

\BFSser{} has both fair \disj{} and fair \conj{}. Our implementation is based 
on \DFSf{} (not \DFSi{}). To implement \BFSser{} based on \DFSf{}, 
we need \texttt{append-map$^\infty_{fair}$} in addition to 
\texttt{append$^\infty_{fair}$}. 
The only difference between \texttt{append-map$^\infty$} and 
\texttt{append-map$^\infty_{fair}$} is that the latter calls 
\texttt{append$^\infty_{fair}$} instead of \texttt{append$^\infty$}.

The implementation is improvable in two ways. First, as mentioned in 
\autoref{sec:fairconj}, \BFSser{} puts answers in bags and answers of the same 
cost are in the 
same bag. In this implementation, however, it is unclear where this information 
is recorded. Second, \texttt{append$^\infty_{fair}$} is extravagant in memory 
usage. It 
makes $O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

%Both \BFSser{} \citet{seres1999algebra} and \BFSopt{} produce answers in 
%increasing order of cost. So it is interesting to see if they are equivalent. 
%We prove so in Coq. The details are in the second subsection.

We prove formally our implementation is equivalent to the original one by 
\citet{seres1999algebra}. The details are in the second subsection.

@@

\NOTE[DPF]{Never refer to a literal section like 2.2. Instead
stick a label where it starts right near the section number and
stick another label where the subsection starts.
Then ref the two labels and place a period between them.}
\NOTE[LKC]{Got it! Fixed}

As mentioned in \autoref{sec:fairconj}, BFS puts answers in bags and 
answers of the 
same cost are in the same bag. The cost
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are all in the same bag), and every thunk indicates an increment 
in cost. It is even more subtle that
\texttt{append$^\infty_{fair}$} and the \texttt{append-map$^\infty_{fair}$} 
respects the cost information. We make these facts more obvious by 
changing the type of search space, modifying related function definitions, 
and introducing a few more functions.

@@

We list functions related to the pure subset in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). We compare these functions with 
\citeauthor{seres1999algebra}'s implementation in our proof. The first three 
functions in Fig.~\ref{BFS-opt} are search space constructors. \texttt{none} 
makes an empty search space; \texttt{unit} makes a space from one answer; and 
\texttt{step} makes a space from a thunk. The remaining functions as before.

@@

\citet{kiselyov2005backtracking} has shown that a \emph{MonadPlus} hides in 
implementations of logic programming system. Our \BFSser{} implementation is 
not an exception: \texttt{append-map$^\infty_{fair}$} is like \texttt{bind}, 
but takes arguments in reversed order; \texttt{none}, \texttt{unit}, and 
\texttt{append$^\infty_{fair}$} correspond to \texttt{mzero}, \texttt{unit}, 
and \texttt{mplus} respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, takes a space \texttt{s$^\infty$} and two 
continuations \texttt{ks} and \texttt{kf}. When \texttt{s$^\infty$} contains 
some 
answers, it calls \texttt{ks} with the first answer and the rest of the space. 
Otherwise, it calls \texttt{kf} with no argument. Here `s' and `f' means 
`succeed' and `fail' respectively. This function is similar to an eliminator of 
search spaces, hence the name. The remaining functions do the same thing as 
before.

\subsection{compare our implementation with the original one}

In this section, we compare the pure subset of our implementation with the 
original one. We focus on the pure subset because the original system is pure.
\citeauthor{seres1999algebra} represent search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate \BFSser{}'s Haskell code into Racket (See supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower than our 
implementation. Details about differences in efficiency are in the table below.

We use Coq to show that the two implementations are 
semantically equivalent, (i.e., \texttt{(run n ? g)} 
produces the same result (See supplements for the formal proof).

@@

\citet{seres1999algebra} describe a breadth-first search 
strategy. We present another implementation. Our implementation is semantically 
equivalent to theirs. But, ours is shorter and performs better in comparison 
with a straightforward translation of their Haskell code.
